---
title: Analysis of the health and economic consequences of storms and other severe
  weather events
author: "treadrightly"
date: "06/20/2015"
output: html_document
---
## Introduction
This document analyses the health and economic consequences of storms and other severe events, using the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data Processing
The following code block downloads the data and loads it for further processing.

```{r cache=TRUE}
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2" ,
              dest = "storm_data.csv.bz2", quiet="TRUE", method="curl")
data <- read.csv("storm_data.csv.bz2", header=TRUE, sep=",")
```

## Results

### Event type most harmful to human health
The first question asked of this report is:  
Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
  
To answer this question, we first identify the columns that provide data about population health. In this report, we process the number of fatalities and number of injuries as the columns describing the impact on population health. We summarize the number of fatalities and number of injuries by event type. We then add the number of fatalities and injuries per event type to arrive at the final number indicating the number of affected people by each event type.

```{r}
fatalities <- aggregate(data$FATALITIES, by=list(data$EVTYPE), FUN=sum)
injuries <- aggregate(data$INJURIES, by=list(data$EVTYPE), FUN=sum)
affected <- data.frame("EVTYPE" = fatalities$Group.1, "AFFECTED" = fatalities$x + injuries$x)
```

To list the event types that are most harmful to population health, we sort the summary dataframe in decreasing order of affected people, and present the top 10.

```{r}
affected <- affected[order(affected$AFFECTED, decreasing=TRUE), ]
head(affected, n=10)
par(las = 1, mar = c(4, 7, 1, 1))
with(affected[1:10, ], barplot(AFFECTED, names.arg=EVTYPE, horiz=TRUE, xlab="Impact on health", ylab="Event Type"))

```

### Event types with greatest economic consequences
The second question asked of this report is:  
Across the United States, which types of events have the greatest economic consequences?  
  
To answer this question, we first identify the columns that provide data about economic consequences. In this report, we process the property damage and crop damage as the columns describing the economic consequences. We summarize the property damage and crop damage by the event type. We then add the property damage and crop damage per event type to arrive at the total damage per event type.

```{r}
propertyDamage <- aggregate(data$PROPDMG, by=list(data$EVTYPE), FUN=sum)
cropDamage <- aggregate(data$CROPDMG, by=list(data$EVTYPE), FUN=sum)
ec <- data.frame("EVTYPE" = propertyDamage$Group.1, "DAMAGE" = propertyDamage$x + cropDamage$x)
```

To list the event types that have the greatest economic consequences, we sort the summary dataframe in decreasing order of damage, and present the top 10.

```{r}
ec <- ec[order(ec$DAMAGE, decreasing=TRUE), ]
head(ec, n=10)
par(las = 1, mar = c(4, 7, 1, 1))
with(ec[1:10, ], barplot(DAMAGE, names.arg = EVTYPE, horiz = TRUE, xlab="Economic Consequences", ylab="Event Type"))
```
